Improving Haskell Theory Exploration

### 2015-12-01 TransferReport/background.tex
Feature extraction

The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions). We also omit several other forms of literal (machine words of various sizes, individual characters, etc.) for brevity, as their treatment is similar to those of strings and numerals. We denote names and literals using quoted strings, e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}.

@@

    factorial :: Nat -> Nat
    factorial    Z  = S Z
    factorial (S n) = mult (S n) (factorial n)

@@

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype and functions for Peano arithmetic; in particular, the \hs{even} and \hs{odd} functions are mutually recursive. The \hs{Bool} type is Haskell's built in representation of Booleans, which can be regarded as \hs{data Bool = True \| False}.}
  \label{fig:haskellexample}
\end{figure}

\begin{figure}
  \begin{subfigure}[plus]{\textwidth}
    \begin{small}
      \underline{\texttt{plus}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Local "y")),
                        (DataAlt "S", ["x"], App (Var (Global "S"))
                                                 (App (App (Var (Global "plus"))
                                                           (Var (Local  "x")))
                                                      (Var (Local "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[mult]{\textwidth}
    \begin{small}
      \underline{\texttt{mult}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Global "Z")),
                        (DataAlt "S", ["x"], App (App (Var (Global "plus"))
                                                      (Var (Local  "y")))
                                                 (App (App (Var (Global "mult"))
                                                           (Var (Local  "x")))
                                                      (Var (Local  "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[factorial]{\textwidth}
    \begin{small}
      \underline{\texttt{factorial}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    App (Var (Global "S"))
                                        (Var (Global "Z"))),
               (DataAlt "S", ["n"], App (App (Var (Global "mult"))
                                             (App (Var (Global "S"))
                                                  (Var (Local  "n"))))
                                        (App (Var (Global "factorial"))
                                             (Var (Local  "n"))))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[odd]{\textwidth}
    \begin{small}
      \underline{\texttt{odd}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "False")),
               (DataAlt "S", ["n"], App (Var (Global "even"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[even]{\textwidth}
    \begin{small}
      \underline{\texttt{even}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "True")),
               (DataAlt "S", ["n"], App (Var (Global "odd"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}. Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions. Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals. The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}. Although the Core is more verbose, we can see similar structure across the definitions (such as \texttt{(App (App (Var (Global $\blank$)) (Var (Local "x"))) (Var (Local "y")))}) which arise from similarities in the Haskell definitions (in this case, \hs{(plus x y)} and \hs{(mult x y)}). This is clearest in the definitions of \hs{odd} and \hs{even}, which are identical except for the global variables.

@@

\qspec{} (and \hspec{}) are also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.

@@

\iffalse TODO: include definitions \fi
\iffalse TODO: make more relevant to our context \fi

>>>>

The sub-set of Core we consider is shown in figure \ref{fig:coresyntax}; compared to the full language \footnote{As of GHC version 7.10.2, the latest at the time of writing.} our major change is to ignore type hints (such as explicit casts, and differences between types/kinds/coercions). For brevity, we also omit several other forms of literal (machine words of various sizes, individual characters, etc.), as their treatment is similar to those of strings and numerals. We will use quoted strings to denote names and literals, e.g. \texttt{Local "foo"}, \texttt{Global "bar"}, \texttt{LitStr "baz"} and \texttt{LitNum "42"}, and require only that they can be compared for equality.

@@

@@

    even :: Nat -> Bool
    even    Z  = True
    even (S n) = odd n
  \end{haskell}
  \caption{A Haskell datatype for Peano numerals with some simple arithmetic functions, including mutually-recursive definitions for \hs{odd} and \hs{even}. \hs{Bool} is Haskell's built in boolean type, which can be regarded as \hs{data Bool = True \| False}.}
  \label{fig:haskellexample}
\end{figure}

\begin{figure}
  \begin{subfigure}[plus]{\textwidth}
    \begin{small}
      \underline{\texttt{plus}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Local "y")),
                        (DataAlt "S", ["x"], App (Var (Global "S"))
                                                 (App (App (Var (Global "plus"))
                                                           (Var (Local  "x")))
                                                      (Var (Local "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[mult]{\textwidth}
    \begin{small}
      \underline{\texttt{mult}}
      \begin{verbatim}
Lam "a" (Lam "y" (Case (Var (Local "a"))
                       "b"
                       [(DataAlt "Z", [],    Var (Global "Z")),
                        (DataAlt "S", ["x"], App (App (Var (Global "plus"))
                                                      (Var (Local  "y")))
                                                 (App (App (Var (Global "mult"))
                                                           (Var (Local  "x")))
                                                      (Var (Local  "y"))))]))
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[odd]{\textwidth}
    \begin{small}
      \underline{\texttt{odd}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "False")),
               (DataAlt "S", ["n"], App (Var (Global "even"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \begin{subfigure}[even]{\textwidth}
    \begin{small}
      \underline{\texttt{even}}
      \begin{verbatim}
Lam "a" (Case (Var (Local "a"))
              "b"
              [(DataAlt "Z", [],    Var (Global "True")),
               (DataAlt "S", ["n"], App (Var (Global "odd"))
                                        (Var (Local  "n")))])
      \end{verbatim}
    \end{small}
  \end{subfigure}
  \caption{Translations of functions in figure \ref{fig:haskellexample} into the Core syntax of figure \ref{fig:coresyntax}. Notice the introduction of explicit $\lambda$ abstractions (\texttt{Lam}) and the use of \texttt{Case} to represent piecewise definitions. Fresh variables are chosen arbitrarily as \texttt{"a"}, \texttt{"b"}, etc.}
  \label{fig:coreexample}
\end{figure}

Figure \ref{fig:haskellexample} shows some simple Haskell function definitions, along with a custom datatype for Peano numerals. The translation to our Core syntax is routine, and shown in figure \ref{fig:coreexample}. Although the Core is more verbose, we can see that similar structure in the Haskell definitions gives rise to similar structure in the Core; for example, the definitions of \hs{odd} and \hs{even} are identical in both languages, except for the global variables. This correspondence allows us to analyse

@@

\qspec{} (and \hspec{}) is also compatible with Haskell's existing testing infrastructure, such that an invocation of \texttt{cabal test} can run these tools alongside more traditional QA tools like \qcheck{}, \textsc{HUnit} and \textsc{Criterion}.

@@

One of our main contributions is a \emph{feature extraction} algorithm for Core expressions. Feature extraction is the conversion of ``raw'' input data, such as audio, images or (in our case) expressions, into a form more suited to machine learning algorithms. By pre-processing our data in this way, we can re-use the same ``off-the-shelf'' machine learning algorithms in a variety of domains.

We use a standard representation of features known as a \emph{feature vector}: a sequence of numbers $(x_1, \dots, x_n)$ where $x_i \in \mathbb{R}$. \footnote{In fact, we only use $\mathbb{R}$ for notational and analytic convenience. Practical implementations will use an approximate format such as IEEE 754 floating point numbers.} For learning purposes this has two important advantages over our raw expressions:

\begin{itemize}
  \item All of our feature vectors will be the same size, i.e. they will all have length (or \emph{dimension}) $n$. This makes element-wise comparison between arbitrary feature vectors $x$ and $y$ trivial, as each element $x_i$ will have a corresponding element $y_i$. For raw expressions this is not so straightforward, as their nesting may give rise to very different shapes.
  \item Unlike our discrete expressions, we can continuously transform one feature vector into another. This enables many powerful ML algorithms to be used, such as those based on \emph{gradient descent} or, in our case, taking the arithmetic mean.
\end{itemize}
